---
title: "Most Harmful Weather Events"
author: "Miguel Briones"
date: "December 25, 2016"
output: html_document
---

Synopsis
---
The most harmful weather events across the United States, in relation to population health, are tornadoes, excessive heat, and flash floods.

Weather events with the greatest economic consequences, across the United States, are tornadoes, thunderstorm winds, and flash floods. 

The raw data is taken from the [National Weather Service Instruction 10-1605] [1]. The database begins in the year 1950 and ends in November of 2011. Injuries, fatalities, and property damage (in dollars) are totalled over that time. 

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf "National Weather Service Instruction 10-1605"

Data Processing
---------------
First, we load the data:
```{r cache=TRUE}
storm.data = read.csv(bzfile("repdata%2Fdata%2FStormData.csv.bz2"), header=TRUE)
```

We remove some of the columns:
```{r}
reduced.storm.data <- 
        storm.data[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG")]
```

Then we normalize the event name:
```{r}
reduced.storm.data$EVTYPE <- 
    gsub("^HEAT$", "EXCESSIVE HEAT", reduced.storm.data$EVTYPE)
reduced.storm.data$EVTYPE <- 
    gsub("^TSTM WIND$", "THUNDERSTORM WIND", reduced.storm.data$EVTYPE)
reduced.storm.data$EVTYPE <- 
    gsub("^THUNDERSTORM WIND$", "THUNDERSTORM WINDS", reduced.storm.data$EVTYPE)
```


Now we aggregate the data on fatalities and find the top 10 causes of deaths:
```{r}
agg.fatalities.data=aggregate(reduced.storm.data$FATALITIES,by=list(reduced.storm.data$EVTYPE), FUN=sum, na.rm=TRUE)
colnames(agg.fatalities.data) = c("event.type", "fatality.total")
fatalities.sorted=agg.fatalities.data[order(-agg.fatalities.data$fatality.total),]
top.fatalities=fatalities.sorted[1:10,]
top.fatalities$event.type=factor(top.fatalities$event.type, levels=top.fatalities$event.type, ordered=TRUE)
```

We then do the same for injuries:
```{r}
agg.injuries.data <-
        aggregate(
                reduced.storm.data$INJURIES, 
                by=list(reduced.storm.data$EVTYPE), FUN=sum, na.rm=TRUE)
colnames(agg.injuries.data) = c("event.type", "injury.total")
injuries.sorted <- agg.injuries.data[order(-agg.injuries.data$injury.total),] 
top.injuries <- injuries.sorted[1:10,]
top.injuries$event.type <- 
    factor(
        top.injuries$event.type, levels=top.injuries$event.type, 
        ordered=TRUE)
```

Finally we do the same for property damage:
```{r}
agg.prop.dmg.data <-
        aggregate(
                reduced.storm.data$PROPDMG, 
                by=list(reduced.storm.data$EVTYPE), FUN=sum, na.rm=TRUE)
colnames(agg.prop.dmg.data) = c("event.type", "prop.dmg.total")
prop.dmg.sorted <- agg.prop.dmg.data[order(-agg.prop.dmg.data$prop.dmg.total),] 
top.prop.dmg <- prop.dmg.sorted[1:10,]
top.prop.dmg$event.type <- 
    factor(
        top.prop.dmg$event.type, levels=top.prop.dmg$event.type, 
        ordered=TRUE)
```
Results
-------
First we graph the top 10 causes of fatalities:
```{r}
library(ggplot2)
ggplot(data=top.fatalities, aes(x=event.type, y=fatality.total)) + geom_bar(stat="identity") + xlab("Event type") + ylab("Total Fatalities") + ggtitle("Fatalities By Event Type") + theme(axis.text.x = element_text(angle = 45, hjust=1))
```
We do the same for injuries:
```{r}
ggplot(data=top.injuries, aes(x=event.type, y=injury.total)) + 
    geom_bar(stat="identity") + xlab("Event type") + ylab("Total injuries") + 
    ggtitle("Injuries By Event Type") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Then, we do so for property damage:
```{r}
ggplot(data=top.prop.dmg, aes(x=event.type, y=prop.dmg.total)) + 
    geom_bar(stat="identity") + xlab("Event type") + 
    ylab("Total property damage") +  ggtitle("Property Damage By Event Type") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```